Wald test to evaluate regression and dispersion parameters in multivariate covariance generalized linear models

### 2022-07-27 5_artigos/statistic-in-medicine/draft/1-DRAFT1/draft-sm-lineu.tex
organiza drafts

Clinical trials are common in medical research. This type of study usually generates data with multiple non-Gaussian responses and time-dependent observations. The analysis of data from these studies requires statistical modeling techniques that take these characteristics into account. Our proposal is to use the Wald test to perform general hypothesis tests, ANOVAs, MANOVAs and multiple comparison tests on regression and dispersion parameters of multivariate covariance generalized linear models (McGLMs). McGLMs provide a general statistical modeling framework for normal and non-normal multivariate data analysis along with a wide range of  correlation structures. The test was verified based on simulation studies and the results showed that the further the hypothesis tested is from the true values of the parameters, the greater the percentage of rejection of the null hypothesis. The lowest rejection percentages were observed when the hypothesis tested corresponded to the real values of the parameters. Furthermore, the larger the sample size, the greater the power of the test. The proposal is motivated by the analysis of a clinical trial that aims to evaluate the effect of the use of probiotics in the control of addiction and binge eating disorder in patients undergoing bariatric surgery. The subjects were separated into two groups (placebo and treatment) and evaluated at three different times. The results indicate that addiction and binge eating disorder reduce over time, but there is no difference between groups at each time point.

@@

According to \cite{meinert1986clinical} "a clinical trial is a planned experiment designed to assess the efficacy of a treatment in man by comparing the outcomes in a group of patients treated with the test with those observed in a comparable group of patients receiving a control treatment where patients in both groups are enrolled, treated, and followed over the same time period".

@@

Often, in medical studies that use clinical trials, multiple outcomes are taken for a group of patients, for example: \cite{kangovi2018effect} evaluated the effect of community health worker support on clinical outcomes of low-income patients; \cite{song2019effect} studied the effect of a workplace wellness program on employee health and economic outcomes; \cite{thyregod2019five} presents the results of a five-year clinical and echocardiographic outcomes from the nordic aortic valve intervention (NOTION); \cite{schmitz2019effect} evalueated the effect of Home-Based Exercise and Weight Loss Programs on Breast Cancer–Related Lymphedema Outcomes; \cite{duma2019characterization} tested the association between certain exclusion criteria and trial characteristics. Consequently, the joint analysis of multiple outcomes studies has been of increased interest in the medical and statistical literature.

@@

In this study, a set of subjects was divided into two groups: one of them received a placebo and the other received the treatment. In addition, individuals were evaluated at 3 different times: the first evaluation was performed approximately 10 days before surgery; follow-up assessments were performed approximately three months and one year postoperatively.

@@

It is important to note that this is a multivariate problem (contains two response variables), with non-Gaussian responses and that the observations that make up the data set cannot be considered independent, considering that measurements taken on the same individual at different times tend to be correlated. Therefore, it is a problem that traditional modeling techniques would be difficult to apply. It is necessary to use some methodology that supports the requirements of the problem.

@@

When working with regression models, a common interest is to verify if the absence of a certain explanatory variable of the model would generate a loss in the adjustment. Thus, a conjecture of interest is to assess whether there is sufficient evidence in the data to state that a given explanatory variable has no effect on the response. This is done through hypothesis testing. In this context, three hypothesis tests are common: the likelihood ratio test, the Wald test, and the Lagrange multiplier test, also known as the score test. \cite{engle} describes the general formulation of the three tests. All of them are based on the likelihood function of the models.

@@

In the case of traditional linear models, there are techniques such as analysis of variance (ANOVA), initially proposed by \cite{anova_fisher}. According to \cite{anova1}, ANOVA is one of the most widely used statistical methods to test hypotheses and is present in virtually all introductory statistics materials. The objective of the technique is to evaluate the effect of each of the explanatory variables on the response. This is done by comparing models with and without each of the explanatory variables. Therefore, this procedure makes it possible to assess whether the removal of each of the variables generates a significantly worse model when compared to the model with the variable. For the multivariate case, the technique of analysis of variance (ANOVA) is extended to the multivariate analysis of variance \citep{manova}, MANOVA. And among the multivariate hypothesis tests already discussed in the literature, Wilk's \cite{wilks} $\lambda$, Hotelling-Lawley trace \cite{lawley,hotelling}, Pillai trace \cite{pillai} and largest root of Roy \cite{roy}.

@@

Complementary to ANOVAs and MANOVAs are multiple comparison tests. Such procedures are used when the analysis of variance points to the existence of a significant effect of the parameters associated with a categorical variable, that is, there is at least one significant difference between the levels of a factor. Thus, the multiple comparison test is another procedure based on hypothesis testing, used to determine where these differences are. For example, suppose there is a three-level categorical variable $X$ in the model: A, B, and C. The analysis of variance will show if there is an effect of the variable $X$ in the model, that is, if the response values are associated with the levels of $X$, however this result will not tell us whether the response values differ from A to B, or from A to C, or whether B differs from C. To detect such differences, multiple comparison tests are used. Among the tests discussed in the literature are the Dunnett, Tukey, student's t test (LSD), Scott-Knott test, among others. \cite{hsu1996multiple} discusses various procedures for the purpose of multiple comparisons. \cite{bretz2008multiple} presents procedures for multiple comparisons in linear models.

@@

However, for regression problems with multiple non-Gaussian responses, there are few alternatives for hypothesis testing. Therefore, as it is a flexible class of models with high power of application to practical problems, our general objective is the development of hypothesis tests for McGLMs. We have the following specific objectives: to propose the use of the Wald test to carry out tests of general hypotheses on regression and dispersion parameters of McGLMs, also enabling the generation of tables of analysis of variance, multivariate analysis of variance and multiple comparisons tests for regression models with multiple non-Gaussian responses. Another objective is to evaluate the properties of the proposed tests based on simulation studies and to motivate the application potential of the discussed methodologies based on the application to the aforementioned datasets. In the problem, traditional modeling techniques would be difficult to apply, however it is a problem of possible analysis via McGLM and hypothesis tests can be used to evaluate the effect of the interaction between time and use of the probiotic on addiction and disorder of binge eating.

@@

We present the data set in \autoref{sec2}. In \autoref{sec3} we present the review of the general structure and estimation of the parameters of a McGLM, based on the ideas of \cite{Bonat16}. In \autoref{sec4} our proposal is presented with the details of the Wald test to evaluate assumptions about parameters of a McGLM. In \autoref{sec5} we present the results of the performance evaluation of the proposed test based on a simulation study. In the \autoref{sec6} we apply the model to the data and present the main results of the analysis of data from the clinical study that aims to evaluate the use of probiotics in the control of addiction and binge eating disorder in patients undergoing bariatric surgery. Finally, the main results are discussed in \autoref{sec7}, including some directions for future investigations.

@@

The evaluation of binge eating was based on the binge eating scale (BES), one of the most used tools to measure binge eating in research, with numerous studies proving its effectiveness, translated into Portuguese and validated for individuals with obesity and submitted to bariatric surgery. This is a 16-item Likert scale questionnaire, prepared according to the Diagnostic and Statistical Manual of Mental Disorders (3rd edition) \cite{spitzer1980diagnostic} by \cite{gormally1982assessment}. The individuals were instructed to select the option that best represented their answer and the final score was the sum of the points for each item, this score ranging from 0 to 46.

To assess food addiction, the Food Addiction Scale (YFAS) was used, a questionnaire that seeks to detect symptoms of addictive eating behaviors. YFAS was based on substance dependence criteria from Diagnostic and Statistical Manual IV – Proofreading (DSM-IV-TR) \cite{segal2010diagnostic} and endorsed for highly processed foods. This questionnaire was developed by \cite{gearhardt2009preliminary} and is a combination of 25 Likert scale options and the evaluation option used was the number of addiction symptoms.

@@

\begin{itemize}
  \item id: individual identifier variable.
  \item moment: moment identifier variable (T0, T1, T2).
  \item group: group identifier variable (placebo, probiotic).
  \item YFAS: proportion of symptoms that characterize addiction.
  \item BES: proporção de escore de compulsão.
\end{itemize}

@@

The graphical analysis presented in \autoref{fig1} shows, in (a) and (d) that both variables of interest present considerable asymmetry to the right. The boxplots of metrics as a function of group presented in (b) and (e) show significant differences between the placebo and probiotic groups for both responses. The boxplots of the metrics as a function of the evaluation moments, presented in (c) and (f), show that for both metrics the values were higher at T0, with a considerable reduction at T1. When comparing T1 and T2, for YFAS it seems that there is a slight increase in the last evaluation; for BES, T1 and T2 do not seem to differ.

@@

Consider $\boldsymbol{Y}_{N \times R} = \left \{ \boldsymbol{Y}_1, \dots, \boldsymbol{Y}_R \right \}$ a matrix of response variables and $\boldsymbol{ M}_{N \times R} = \left \{ \boldsymbol{\mu}_1, \dots, \boldsymbol{\mu}_R \right \}$ a matrix of expected values. The variance and covariance matrix for each response $r$, $r = 1,..., R$ is denoted by $\Sigma_r$, has dimension $NxN$. In addition, a correlation matrix $\Sigma_b$, of order $R \times R$, which describes the correlation between the response variables, is required.

The McGLMs \citep{Bonat16}are defined by:

@@

\noindent where the $g_r()$ functions are the traditional link functions; $\boldsymbol{X}_r$ denotes a design matrix $N \times k_r$; $\boldsymbol{\beta}_r$ denotes a vector $k_r \times 1$ of regression parameters. $\boldsymbol{\Sigma}_R \overset{G} \otimes \boldsymbol{\Sigma}_b = \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1, \ldots, \tilde{\boldsymbol{ \Sigma}}_R) (\boldsymbol{\Sigma}_b \otimes \boldsymbol{I}) \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1^\top, \ldots, \tilde{\ boldsymbol{\Sigma}}_R^\top)$ is the generalized Kronecker product  \cite{martinez13}, the matrix $\tilde{\boldsymbol{\Sigma}}_r$ denotes the lower triangular matrix of the Cholesky decomposition of the matrix ${\boldsymbol{\Sigma}}_r$. The operator $\mathrm{Bdiag()}$ denotes the block-diagonal matrix and $\boldsymbol{I}$ is an identity matrix $N \times N$.

@@

The dispersion matrix $\boldsymbol{\Omega({\tau})}$ describes the part of the covariance within each response variable that does not depend on the mean structure, ie, the correlation structure between the sample observations. Based on the ideas of \cite{Anderson73} and \cite{Pourahmadi00}, \cite{Bonat16} proposed to model the dispersion matrix through a matrix linear predictor combined with a covariance link function given by:

@@

Some possible covariance link functions are identity, inverse and exponential-matrix. The specification of the covariance link function is discussed by \cite{Pinheiro96} and it is possible to select combinations of matrices to obtain the most known models in the literature for longitudinal data, time series, spatial and spatiotemporal data. Further details are discussed by \cite{Demidenko13}.

@@

McGLMs are fitted are fitted based on the estimating function approach described in detail by \cite{Bonat16} and \cite{jorg04}. This subsection presents an overview of the algorithm and the asymptotic distribution of estimators based on estimating functions.

@@

To take into account the covariance between the vectors $\boldsymbol{\beta}$ and $\boldsymbol{\lambda}$, \cite{Bonat16} obtained the sensitivity and cross-variability matrices, denoted by $S_{\boldsymbol{ \lambda \beta}}$, $S_{\boldsymbol{\beta \lambda}}$ and $V_{\boldsymbol{\lambda \beta}}$, more details in \cite{Bonat16}. The joint sensitivity and variability matrices of $\psi_{\boldsymbol{\beta}}$ and $\psi_{\boldsymbol{\lambda}}$ are denoted by

@@

To solve the system of equations $\psi_{\boldsymbol{\beta}} = 0$ and $\psi_{\boldsymbol{\lambda}} = 0$, the modified Chaser algorithm, proposed by \cite{jorg04}, which is defined as

@@

This section is dedicated to the presentation of our proposal: the use of the Wald test to evaluate McGLMs parameters. It is worth remembering that in McGLMs there are regression, dispersion, power and correlation parameters and that, in this work, we are interested in the evaluation of regression and dispersion parameters. Each set of parameters has a very relevant practical interpretation in such a way that through the regression parameters it is possible to identify the relevant explanatory variables, through the dispersion parameters it is possible to evaluate the impact of the correlation between units of the data set, through the power parameters it is possible to identify which probability distribution best suits the problem according to the variance function, and through the correlation parameters it is possible to assess the association between responses.

In \autoref{sec5} we show that the parameter estimates of a McGLM follow a normal distribution. Therefore any quadratic form obtained through the estimator or linear combinations (such as the Wald statistic) is also a quadratic form that has a chi-square distribution. Therefore, our proposal is not a simple adaptation, but an asymptotically valid procedure based on the theory of the distribution of quadratic forms. More about the distribution of quadratic forms can be seen in the works of \cite{graybill1957idempotent}, \cite{luther1965decomposition} and \cite{baldessari1967distribution}.

@@

For the purposes of illustrating the types of hypotheses mentioned, consider the situation in which you want to investigate whether a numeric variable $X_1$ has an effect on two response variables, denoted by $Y_1$ and $Y_2$. For this task, a sample with $N$ observations was collected and for each observation the values of $X_1$, $Y_1$ and $Y_2$ were recorded. Based on the collected data, a bivariate McGLM was fitted, with a predictor given by:

@@

In this scenario, questions of interest could be: is there an effect of the variable $X_1$ only on the first response? Or just about the second response? Is it possible that the variable $X_1$ has an effect on both responses at the same time? Is it possible that the effect of the variable is the same for both responses? All these questions can be answered by testing hypotheses on the model parameters and specified using \autoref{eq:hipoteses_wald}. In the next subsections, we present the elements necessary to conduct each test.

@@

Consider the first type of hypothesis: interest in evaluating whether there is an effect of the variable $X_1$ only on the first response. The hypothesis can be written as follows:

@@

\item $\boldsymbol{c}$ = $\begin{bmatrix} 0 \end{bmatrix}$, is the value under the null hypothesis. 
\end{itemize}

@@

\item $\boldsymbol{c} = \begin{bmatrix} 0 \\ 0 \end{bmatrix}$, is the value under the null hypothesis.

@@

Suppose this time that the hypothesis of interest does not involve testing whether the parameter value is equal to a postulated value but checking whether, in the case of this generic problem, the effect of the variable $X_1$ is the same for both responses. In this situation, we would form a hypothesis of equality between the parameters or, in other words, if the difference of the effects is null:

@@

\item $\boldsymbol{c}$ = $\begin{bmatrix} 0 \end{bmatrix}$, is the value under the null hypothesis.

@@

As there is only one hypothesis, the matrix $\boldsymbol{L}$ has only one row. For the matrix $\boldsymbol{L}$ to be correctly specified in the case of an equality hypothesis, it is necessary to put 1 in the column referring to one parameter, and -1 in the column referring to the other parameter, in such a way that the product $\boldsymbol {L}\boldsymbol{\theta^{*}}$ represents the hypothesis of interest initially postulated, in this case the product $\boldsymbol{L}\boldsymbol{\theta^{*}}$ generates exactly the same hypothesis specified in \autoref{eq:ex3} and the asymptotic distribution of the test is $\chi^2_1$.

@@

\autoref{eq:pred_ex} describes a generic bivariate model. It is important to note that in this example both responses are subject to the same predictor. In practice, when it comes to McGLMs, different predictors can be specified between response variables. Thus, what was exposed in \autoref{sec:ex2} is useful for any case in which there is interest in testing hypotheses about more than one model parameter in the same or in different responses, regardless of the predictors between responses.

However, in cases where the responses are subject to identical predictors and the assumptions about the parameters do not change from response to response, an alternative specification of the procedure is to use the Kronecker product to test the same hypothesis on multiple responses as used in \cite{plastica}.

Suppose that, in this example, the hypotheses of interest are still written as in \autoref{eq:ex2}. However, as this is a bivariate model with the same predictor for the two responses, the hypothesis of interest is the same between responses and involves only regression parameters, it is convenient to write the matrix $\boldsymbol{L}$ as the Kronecker product from two matrices: a matrix $\boldsymbol{G}$ and a $\boldsymbol{F}$, that is, $\boldsymbol{L}$ = $\boldsymbol{G} \otimes \boldsymbol{F}$. In this way, the matrix $\boldsymbol{G}$ has dimension $R \times R$ and specifies the hypotheses regarding the responses, whereas the matrix $\boldsymbol{F}$ specifies the hypotheses between variables and has dimension ${s} ' \times {h}'$, where ${s}'$ is the number of linear constraints, that is, the number of parameters tested for a single response, and ${h}'$ is the total number of coefficients of regression or dispersion of the response. Therefore, the array $\boldsymbol{L}$ has dimension (${s}'R \times h$).

In general, the matrix $\boldsymbol{G}$ is an identity matrix with a dimension equal to the number of responses analyzed in the model. Whereas the matrix $\boldsymbol{F}$ is equivalent to a matrix $\boldsymbol{L}$ if there was only a single response in the model and only regression or dispersion parameters. We use the Kronecker product of these two matrices to ensure that the hypothesis described in the $\boldsymbol{F}$ matrix is tested on the $R$ model responses.

@@

\item $\boldsymbol{c} = \begin{bmatrix} 0 \\ 0 \end{bmatrix}$, is the value under the null hypothesis.

@@

Thus, the product $\boldsymbol{L}\boldsymbol{\beta}$ represents the initially postulated hypothesis of interest. In this case, the asymptotic distribution of the test is $\chi^2_2$. The procedure is easily generalized when there is interest in evaluating a hypothesis about the dispersion parameters and this specification is quite convenient for generating analysis of variance tables.

@@

For the purposes of illustrating the tests performed by each type of analysis of variance proposed, consider the situation in which we want to investigate whether two numeric variables denoted by $X_1$ and $X_2$ have an effect on two response variables denoted by $Y_1$ and $Y_2$. For this task, a sample with $N$ observations was collected and for each observation the values of $X_1$, $X_2$, $Y_1$ and $Y_2$ were recorded. Based on the collected data, a bivariate model was fitted, with a predictor given by:

@@

Note that, in the lines that seek to understand the effect of $X_1$ and $X_2$, the interaction is also evaluated, as all parameters involving that variable are removed from the model.

>>>>

Clinical trials are common in medical research. This type of study usually generates data with multiple non-Gaussian responses and time-dependent observations. The analysis of data from these studies requires statistical modeling techniques that take these characteristics into account. Our proposal is to use the Wald test to perform general hypothesis tests, ANOVAs, MANOVAs and multiple comparison tests on regression and dispersion parameters of multivariate covariance generalized linear models (McGLMs). McGLMs provide a general statistical modeling framework for normal and non-normal multivariate data analysis along with a wide range of  correlation structures. The proposed test was verified based on simulation studies and the results showed that the further the hypothesis tested is from the true values of the parameters, the greater the percentage of rejection of the null hypothesis. The lowest rejection percentages were observed when the hypothesis tested corresponded to the real values of the parameters. Furthermore, the larger the sample size, the greater the power of the test. The proposal is motivated by the analysis of a clinical trial that aims to evaluate the effect of the use of probiotics in the control of addiction and binge eating disorder in patients undergoing bariatric surgery. The subjects were separated into two groups (placebo and treatment) and evaluated at three different times. The results indicate that addiction and binge eating disorder reduce over time, but there is no difference between groups at each time point.

@@

According to Meinert and Tonascia \cite{meinert1986clinical} "a clinical trial is a planned experiment designed to assess the efficacy of a treatment in man by comparing the outcomes in a group of patients treated with the test with those observed in a comparable group of patients receiving a control treatment where patients in both groups are enrolled, treated, and followed over the same time period".

@@

Often, in medical studies that use clinical trials, multiple outcomes are taken for a group of patients, for example: Kangovi et al \cite{kangovi2018effect} evaluated the effect of community health worker support on clinical outcomes of low-income patients; Song and Baicker \cite{song2019effect} studied the effect of a workplace wellness program on employee health and economic outcomes; Thyregod et al \cite{thyregod2019five} presents the results of a five-year clinical and echocardiographic outcomes from the nordic aortic valve intervention (NOTION); Schmitz et al \cite{schmitz2019effect} evalueated the effect of Home-Based Exercise and Weight Loss Programs on Breast Cancer–Related Lymphedema Outcomes; Duma et al \cite{duma2019characterization} tested the association between certain exclusion criteria and trial characteristics. Consequently, the joint analysis of multiple outcomes studies has been of increased interest in the medical and statistical literature.

@@

In this study, a set of subjects was divided into two groups: one of them received a placebo and the other received the treatment. In addition, individuals were evaluated at 3 different time points: the first evaluation was performed approximately 10 days before surgery; follow-up assessments were performed approximately three months and one year postoperatively.

@@

It is important to note that this is a multivariate problem (contains two response variables), with non-Gaussian responses and that the observations that make up the data set cannot be considered independent, considering that measurements taken on the same individual at different time points tend to be correlated. Therefore, it is a problem that traditional modeling techniques would be difficult to apply. It is necessary to use some methodology that supports the requirements of the problem.

@@

When working with regression models, a common interest is to verify if the absence of a certain explanatory variable of the model would generate a loss in the adjustment. Thus, a conjecture of interest is to assess whether there is sufficient evidence in the data to state that a given explanatory variable has no effect on the response. This is done through hypothesis testing. In this context, three hypothesis tests are common: the likelihood ratio test, the Wald test, and the Lagrange multiplier test, also known as the score test. Engle \cite{engle} describes the general formulation of the three tests. All of them are based on the likelihood function of the models.

@@

In the case of traditional linear models, there are techniques such as analysis of variance (ANOVA), initially proposed by Fisher and Mackenzie \cite{anova_fisher}. According to St, Wold et al \cite{anova1}, ANOVA is one of the most widely used statistical methods to test hypotheses and it is present in virtually all introductory statistics materials. The objective of the technique is to evaluate the effect of each explanatory variables on the response. This is done by comparing models with and without each of the explanatory variables. Therefore, this procedure makes it possible to assess whether the removal of each of the variables generates a significantly worse model when compared to the model with the variable. For the multivariate case, the technique of analysis of variance (ANOVA) is extended to the multivariate analysis of variance \citep{manova}, MANOVA. And among the multivariate hypothesis tests already discussed in the literature, we have the Wilk's \cite{wilks} $\lambda$, Hotelling-Lawley trace \cite{lawley,hotelling}, Pillai trace \cite{pillai} and largest root of Roy \cite{roy}.

@@

Complementary to ANOVAs and MANOVAs are multiple comparison tests. Such procedures are used when the analysis of variance points to the existence of a significant effect of the parameters associated with a categorical variable, that is, there is at least one significant difference between the levels of a factor. Thus, the multiple comparison test is another procedure based on hypothesis testing, used to determine where these differences are. For example, suppose there is a three-level categorical variable $X$ in the model that assumes the levels: A, B, or C. The analysis of variance will show if there is an effect of the variable $X$ in the model, that is, if the response values are associated with the levels of $X$, however this result will not tell us whether the response values differ from A to B, or from A to C, or whether B differs from C. To detect such differences, multiple comparison tests are used. Among the tests discussed in the literature are the Dunnett, Tukey, student's t test (LSD), Scott-Knott test, among others. Hsu \cite{hsu1996multiple} discusses various procedures for the purpose of multiple comparisons. Bretz, Hothorn, and Westfall \cite{bretz2008multiple} presents procedures for multiple comparisons in linear models.

@@

However, for regression problems with multiple non-Gaussian responses, there are few alternatives for hypothesis testing. Therefore, as it is a flexible class of models with high power of application to practical problems, our general objective is the development of hypothesis tests for McGLMs. We have the following specific objectives: to propose the use of the Wald test to carry out tests of general hypotheses on regression and dispersion parameters of McGLMs, also enabling the generation of tables of analysis of variance, multivariate analysis of variance and multiple comparisons tests for regression models with multiple non-Gaussian responses. Another objective is to evaluate the properties of the proposed tests based on simulation studies and to motivate the application potential of the discussed methodologies based on the analysis of the aforementioned datasets. In the problem, traditional modeling techniques would be difficult to apply, however it is a problem of possible analysis via McGLM and hypothesis tests can be used to evaluate the effect of the interaction between time and use of the probiotic on addiction and disorder of binge eating.

@@

We present the data set in \autoref{sec2}. In \autoref{sec3} we present the review of the general structure and estimation of the parameters of a McGLM, based on the ideas of Bonat and Jørgensen \cite{Bonat16}. In \autoref{sec4} our proposal is presented with the details of the Wald test to evaluate assumptions about parameters of a McGLM. In \autoref{sec5} we present the results of the performance evaluation of the proposed test based on a simulation study. In the \autoref{sec6} we apply the model to the data and present the main results of the analysis of data from the clinical study that aims to evaluate the use of probiotics in the control of addiction and binge eating disorder in patients undergoing bariatric surgery. Finally, the main results are discussed in \autoref{sec7}, including some directions for future investigations.

@@

The evaluation of binge eating was based on the binge eating scale (BES), one of the most used tools to measure binge eating in research, with numerous studies proving its effectiveness, translated into Portuguese and validated for individuals with obesity and submitted to bariatric surgery. This is a 16-item Likert scale questionnaire, prepared according to the Diagnostic and Statistical Manual of Mental Disorders (3rd edition) \cite{spitzer1980diagnostic} by Gormally et al \cite{gormally1982assessment}. The individuals were instructed to select the option that best represented their answer and the final score was the sum of the points for each item, this score ranging from 0 to 46.

To assess food addiction, the Food Addiction Scale (YFAS) was used, a questionnaire that seeks to detect symptoms of addictive eating behaviors. YFAS was based on substance dependence criteria from Diagnostic and Statistical Manual IV – Proofreading (DSM-IV-TR) \cite{segal2010diagnostic} and endorsed for highly processed foods. This questionnaire was developed by Gearhardt et al \cite{gearhardt2009preliminary} and is a combination of 25 Likert scale options and the evaluation option used was the number of addiction symptoms.

@@

\begin{itemize}
  \item id: individual identifier variable.
  \item moment: moment identifier variable (T0, T1, T2).
  \item group: group identifier variable (placebo, probiotic).
  \item YFAS: proportion of symptoms that characterize addiction.
  \item BES: BES score transformed to unit scale.
\end{itemize}

@@

The graphical analysis presented in \autoref{fig1} shows, in (a) and (d) that both variables of interest present considerable asymmetry to the right. The boxplots of metrics as a function of group presented in (b) and (e) show litthe difference between the placebo and probiotic groups for both responses. The boxplots of the metrics as a function of the evaluation moments, presented in (c) and (f), show that for both metrics the values were higher at T0, with a considerable reduction at T1. When comparing T1 and T2, for YFAS it seems that there is a slight increase in the last evaluation; for BES, T1 and T2 do not seem to differ.

@@

Consider $\boldsymbol{Y}_{N \times R} = \left \{ \boldsymbol{Y}_1, \dots, \boldsymbol{Y}_R \right \}$ a matrix of response variables and $\boldsymbol{ M}_{N \times R} = \left \{ \boldsymbol{\mu}_1, \dots, \boldsymbol{\mu}_R \right \}$ a matrix of expected values. The variance and covariance matrix for each response $r$, $r = 1,..., R$, is denoted by $\Sigma_r$ and has dimension $N \times N$. In addition, consider a correlation matrix $\Sigma_b$, of order $R \times R$, which describes the correlation between the response variables.

The McGLMs \citep{Bonat16} are defined by:

@@

\noindent where the $g_r()$ are the traditional link functions; $\boldsymbol{X}_r$ denotes a design matrix $N \times k_r$; $\boldsymbol{\beta}_r$ denotes a vector $k_r \times 1$ of regression parameters. $\boldsymbol{\Sigma}_R \overset{G} \otimes \boldsymbol{\Sigma}_b = \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1, \ldots, \tilde{\boldsymbol{ \Sigma}}_R) (\boldsymbol{\Sigma}_b \otimes \boldsymbol{I}) \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1^\top, \ldots, \tilde{\boldsymbol{\Sigma}}_R^\top)$ is the generalized Kronecker product  \cite{martinez13}, the matrix $\tilde{\boldsymbol{\Sigma}}_r$ denotes the lower triangular matrix of the Cholesky decomposition of the matrix ${\boldsymbol{\Sigma}}_r$. The operator $\mathrm{Bdiag()}$ denotes the block-diagonal matrix and $\boldsymbol{I}$ is an identity matrix $N \times N$.

@@

The dispersion matrix $\boldsymbol{\Omega({\tau})}$ describes the part of the covariance within each response variable that does not depend on the mean structure, ie, the correlation structure between the sample observations. Based on the ideas of Anderson et al \cite{Anderson73} and Pourahmadi \cite{Pourahmadi00}, Bonat and Jørgensen \cite{Bonat16} proposed to model the dispersion matrix through a matrix linear predictor combined with a covariance link function given by:

@@

Some possible covariance link functions are identity, inverse and exponential-matrix. The specification of the covariance link function is discussed by Pinheiro and Bates \cite{Pinheiro96} and it is possible to select combinations of matrices to obtain the most known models in the literature for longitudinal data, time series, spatial and spatiotemporal data. Further details are discussed by Demidenko \cite{Demidenko13}.

@@

McGLMs are fitted based on the estimating function approach described in detail by Bonat and Jørgensen \cite{Bonat16} and Jørgensen and Knudsen \cite{jorg04}. This subsection presents an overview of the algorithm and the asymptotic distribution of estimators based on estimating functions.

@@

To take into account the covariance between the vectors $\boldsymbol{\beta}$ and $\boldsymbol{\lambda}$, Bonat and Jørgensen \cite{Bonat16} obtained the sensitivity and cross-variability matrices, denoted by $S_{\boldsymbol{ \lambda \beta}}$, $S_{\boldsymbol{\beta \lambda}}$ and $V_{\boldsymbol{\lambda \beta}}$, more details in Bonat and Jørgensen \cite{Bonat16}. The joint sensitivity and variability matrices of $\psi_{\boldsymbol{\beta}}$ and $\psi_{\boldsymbol{\lambda}}$ are denoted by

@@

To solve the system of equations $\psi_{\boldsymbol{\beta}} = 0$ and $\psi_{\boldsymbol{\lambda}} = 0$, Jørgensen and Knudsen \cite{jorg04} proposed the modified Chaser algorithm, which is defined as

@@

This section is dedicated to the presentation of our proposal: the use of the Wald test to evaluate McGLMs parameters. It is worth remembering that in McGLMs there are regression, dispersion, power and correlation parameters. Each set of parameters has a very relevant practical interpretation in such a way that through the regression parameters it is possible to identify the relevant explanatory variables, through the dispersion parameters it is possible to evaluate the impact of the correlation between units of the data set, through the power parameters it is possible to identify which probability distribution best suits the problem according to the variance function, and through the correlation parameters it is possible to assess the association between responses. In this work, we are interested in methods for evaluation of regression and dispersion parameters.

@@

For the purposes of illustrating the types of hypotheses mentioned, consider the situation in which you want to investigate whether a numeric variable $X_1$ has an effect on two response variables, denoted by $Y_1$ and $Y_2$. For this task, consider that a sample with $N$ observations was collected and for each observation the values of $X_1$, $Y_1$ and $Y_2$ were recorded. Based on the collected data, a bivariate McGLM was fitted, with a predictor given by:

@@

In this scenario, questions of interest could be: is there an effect of the variable $X_1$ on only one of the responses? Is it possible that the variable $X_1$ has an effect on both responses at the same time? Is it possible that the effect of the variable is the same for both responses? All these questions can be answered by testing hypotheses on the model parameters and specified using \autoref{eq:hipoteses_wald}. In the next subsections, we present the elements necessary to conduct each test.

@@

Let's consider the simplest type of hypothesis to test: a hypothesis about a single parameter. Suppose our interest is evaluate whether there is an effect of the variable $X_1$ only on the first response. The hypothesis can be written as follows:

@@

\item $\boldsymbol{c}$ = $\begin{bmatrix} 0 \end{bmatrix}.$

@@

\item $\boldsymbol{c} = \begin{bmatrix} 0 \\ 0 \end{bmatrix}.$

@@

Suppose this time that the hypothesis of interest does not involve testing whether the parameter value is equal to a postulated value, but checking whether the effect of the variable $X_1$ is the same for both responses. In this situation, we would form a hypothesis of equality between the parameters or, in other words, if the difference of the effects is null:

@@

\item $\boldsymbol{c}$ = $\begin{bmatrix} 0 \end{bmatrix}.$

@@

As there is only one hypothesis, the matrix $\boldsymbol{L}$ has only one row. For the matrix $\boldsymbol{L}$ to be correctly specified in the case of an equality hypothesis, it is necessary to put 1 in the column referring to one parameter, and -1 in the column referring to the other parameter, in such a way that the product $\boldsymbol {L}\boldsymbol{\theta^{*}}$ represents the hypothesis of interest initially postulated. In this case the product $\boldsymbol{L}\boldsymbol{\theta^{*}}$ generates exactly the same hypothesis specified in \autoref{eq:ex3} and the asymptotic distribution of the test is $\chi^2_1$.

@@

The \autoref{eq:pred_ex} describes a generic bivariate model. It is important to note that in this example both responses are subject to the same predictor. In practice, when it comes to McGLMs, different predictors can be specified between response variables. Thus, what was exposed in \autoref{sec:ex2} is useful for any case in which there is interest in testing hypotheses about more than one model parameter in the same or in different responses, regardless of the predictors between responses.

However, in cases where the responses are subject to identical predictors and the assumptions about the parameters do not change from response to response, an alternative specification of the procedure is to use the Kronecker product to test the same hypothesis on multiple responses as used in Bonat et al \cite{plastica}.

Suppose that, in this example, the hypotheses of interest are still written as in \autoref{eq:ex2}. However, as this is a bivariate model with the same predictor for the two responses and the hypothesis of interest are the same between responses and involves only regression parameters, it is convenient to write the matrix $\boldsymbol{L}$ as the Kronecker product from two matrices: a matrix $\boldsymbol{G}$ and a $\boldsymbol{F}$, that is, $\boldsymbol{L}$ = $\boldsymbol{G} \otimes \boldsymbol{F}$. In this way, the matrix $\boldsymbol{G}$ has dimension $R \times R$ and specifies the hypotheses regarding the responses, whereas the matrix $\boldsymbol{F}$ specifies the hypotheses between variables and has dimension ${s} ' \times {h}'$, where ${s}'$ is the number of linear constraints, that is, the number of parameters tested for a single response, and ${h}'$ is the total number of coefficients of regression or dispersion of the response. Therefore, the matrix $\boldsymbol{L}$ has dimension (${s}'R \times h$).

In general, the matrix $\boldsymbol{G}$ is an identity matrix with dimension equal to the number of responses analyzed in the model. Whereas the matrix $\boldsymbol{F}$ is equivalent to a matrix $\boldsymbol{L}$ if there was only a single response in the model and only regression or dispersion parameters. We use the Kronecker product of these two matrices to ensure that the hypothesis described in the $\boldsymbol{F}$ matrix will be tested on the $R$ model responses.

@@

\item $\boldsymbol{c} = \begin{bmatrix} 0 \\ 0 \end{bmatrix}$: vector of values under the null hypothesis.

@@

Thus, the product $\boldsymbol{L}\boldsymbol{\beta}$ represents the initially postulated hypothesis of interest. In this case, the asymptotic distribution of the test is $\chi^2_2$. The procedure is easily generalized when there is interest in evaluating a hypothesis about the dispersion parameters. And this specification is quite convenient for generating analysis of variance tables.

@@

For the purposes of illustrating the tests performed by each type of analysis of variance proposed, consider the situation in which we want to investigate whether two numeric variables denoted by $X_1$ and $X_2$ have an effect on two response variables denoted by $Y_1$ and $Y_2$. For this task, suppose that a sample with $N$ observations was collected and for each observation the values of $X_1$, $X_2$, $Y_1$ and $Y_2$ were recorded. Based on the collected data, a bivariate model was fitted, with a predictor given by:

@@

Note that, in the lines that we seek to understand the effect of $X_1$ and $X_2$, the interaction is also evaluated, as all parameters involving that variable are removed from the model.
